---
title: "Final Project"
author: Madeleine Bannon
author: 
author: 
output: html_document
---
@chris @charles - this is the outline for our writeup. feel free to add and/or edit it. 
#Executive summary
##Goal of the study
##Data
###Data cleanup and extraction
###Predictor variables
##Findings

#Detailed analysis
##Model 1 [to be filled in further]
##Model 2 [to be filled in further]
#Appendix
##Exhibit 1
##Exhibit [##]

@chris @charles - proposed managerial  questions [and analyses]
+ What factors impact airline reliability the most? [logistic regression (lasso?) & random forest]
  + What causes a delay vs. what causes a cancellation?
  + Is there a difference between departure delays and arrival delays?

+ What factors impact the length of a cancellation? [multivariate regression (lasso?) & random forest]

[EDA for below?]
+ Which airlines are the most reliable?
+ Which locations experience the least amount of delays?
+ How does time of year impact flight reliability?
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 4)
if(!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, ggplot2, glmnet, car, data.table)   #add your packages here
```

```{r data upload, include=FALSE}


```

```{r}

```

